Efficient Calculation of Polynomial Features on Sparse Matrices

### 2016-11-04 paper/Final/sparse_poly.tex
typos and minor changes

\usepackage{amssymb}
\usepackage{bm}
\usepackage{clrscode3e}
\usepackage{float}
\usepackage{mathtools}

@@

\author{
  Nystrom, Andrew \\
  \texttt{nystrom@google.com}
  \and
  Hughes, John \\
  \texttt{jfh@cs.brown.edu}
}

@@

\begin{abstract}
We provide an algorithm for polynomial feature expansion that operates directly on and produces a compressed sparse row matrix without any densification.
For a vector of dimension $D$, density $d$, and degree $k$ the algorithm has time and space complexity $O(d^kD^k)$ where $k$ is the polynomial order, which is an improvement by a factor $d^k$ over the standard method.
\end{abstract}

@@

We provide an algorithm that allows CSR matrices to be the input of a polynomial feature expansion without any densification.
The algorithm leverages the CSR format to only compute products of features that result in a nonzero value.
This exploits the sparsity of the data to achieve an improved time complexity of $O(d^kD^k)$ on each vector of the matrix where $k$ is the degree of the expansion, $D$ is the dimensionality, and $d$ is the density.
The standard algorithm has time complexity $O(D^k)$.
Since $0 \le d \le 1$, our algorithm is a significant improvement.
While the algorithm we lay out is uses CSR matrices, it could be slightly modified to operate on other sparse formats.

@@

For simplicity, all polynomial expansions are assumed to be of the second degree.
We do provide an algorithm for third degree expansions and derive the big O time complexity of the general case.

@@

\begin{codebox}
\footnotesize
\Procname{$\proc{Incomplete Second Order CSR Polynomial Expansion Algorithm}(\bm{A})$}
    \li $N \gets$ row count of $\bm{A}$
    \li $\bm{p}^k$ $\gets$ vector of size $N+1$
    \li $\bm{p}^k_0 \gets 0$
    \li $nnz^k \gets 0$
    \li \For $i \gets 0 \To N-1$ \Do
    \li     $i_{start} \gets \bm{p}_i$
    \li     $i_{stop} \gets \bm{p}_{i+1}$
    \li     $\bm{c}_i \gets \bm{c}_{i_{start}:i_{stop}}$
    \li     $nnz^k_i \gets \binom{|\bm{c}_i|}{2}$
    \li     $nnz^k \gets nnz^k + nnz^k_i$
    \li     $\bm{p}^k_{i+1} \gets \bm{p}^k_i + nnz^k_i$
        \End
    \zi     
    \zi \Comment Build up the elements of $\bm{p}^k$, $\bm{c}^k$, and $\bm{d}^k$
    \li $\bm{p}^k$ $\gets$ vector of size $N+1$
    \li $\bm{c}^k$ $\gets$ vector of size $nnz^k$
    \li $\bm{d}^k$ $\gets$ vector of size $nnz^k$
    \li $n \gets 0$
    \li \For $i \gets 0 \To N-1$ \Do
    \li     $i_{start} \gets \bm{p}_i$
    \li     $i_{stop} \gets \bm{p}_{i+1}$
    \li     $\bm{c}_i \gets \bm{c}_{i_{start}:i_{stop}}$
    \li     $\bm{d}_i \gets \bm{d}_{i_{start}:i_{stop}}$
    \li     \For $c_1 \gets 0 \To |\bm{c}_i|-1$ \Do
    \li         \For $c_2 \gets c_1 \To |\bm{c}_i|-1$ \Do
%    \li             $j_0 \gets \bm{c}_{c_0}$
%    \li             $j_1 \gets \bm{c}_{c_1}$
    \li             $\bm{d}^k_{n} \gets \bm{d}_{c_0} \cdot \bm{d}_{c_1}$
    \li             $\bm{c}^k_{n} = ?$ \label{li:set_ck}
    \li             $n \gets n + 1$
                \End
            \End
       	\End
\end{codebox}

@@

\begin{equation}
(j_0, j_1, \dots, j_{k-1}) \rightarrowtail \hspace{-1.9ex} \twoheadrightarrow p_{j_0j_1 \dots i_{k-1}} \in \{0,1,\dots,\binom{D}{k}\} 
\end{equation}

@@

\section{Final CSR Expansion Algorithm}
With the mapping from columns of $\bm{A}$ to a column of $\bm{A}^k$, we can now write the final form of the innermost loop of the algorithm from \ref{sec:final-algo}.
Let the mapping for $k=2$ be denoted $PolyCol^2$.
Then the innermost loop becomes:

@@

The algorithm can be generalized to higher degrees by simply adding more nested loops and using higher order mappings.

@@

\subsection{Empirical}
To demonstrate how our algorithm scales with the density of a matrix, we compare it to the traditional polynomial expansion algorithm in the popular machine library scikit-learn \cite{scikit-learn} in the task of generating second degree polynomial expansions.
Matrices of size $100 \times 5000$ were randomly generated with densities of $0.2$, $0.4$, $0.6$, $0.8$, and $1.0$.
Thirty matrices of each density were randomly generated, and the mean times (gray) of each algorithm were plotted.
The red or blue width around the mean marks the third standard deviation from the mean.

@@

The fact that our algorithm is approximately $6.5$ times faster than the scikit-learn algorithm on $100 \times 5000$ matrices that are entirely dense is likely an language implementation difference.
What matters is that the time of our algorithm increases quadratically with respect to the density in accordance with the big O analysis.

@@

\section{Conclusion}
We have developed an algorithm for performing polynomial feature expansions on CSR matrices that scales polynomially with respect to the density of the matrix.
While the areas within machine learning this work touches is not the flavor of the day, they are workhorses of industry and will likely save much coal from being unnecessarily burnt.

>>>>

\usepackage{amssymb}
\usepackage{bm}
\usepackage{clrscode3e}
\usepackage{float}
\usepackage{footmisc}
\usepackage{mathtools}

@@

\newcommand*\samethanks[1][\value{footnote}]{\footnotemark[#1]}
\renewcommand{\thefootnote}{\fnsymbol{thanks}}

\author{
  Nystrom, Andrew \\
  \texttt{awnystrom@gmail.com}
  \thanks{Now at Google}
  \ 
  \thanks{The authors contributed equally important and fundamental aspects of this work.}
  \and
  Hughes, John \\
  \texttt{jfh@cs.brown.edu}
  \footnotemark[2]
}

@@

\begin{abstract}
We provide an algorithm for polynomial feature expansion that operates directly on and produces a compressed sparse row matrix without any densification.
For a vector of dimension $D$, density $d$, and degree $k$ the algorithm has time complexity $O(d^kD^k)$ where $k$ is the polynomial order, which is an improvement by a factor $d^k$ over the standard method.
\end{abstract}

@@

We provide an algorithm that allows CSR matrices to be the input of a polynomial feature expansion without any densification.
The algorithm leverages the CSR format to only compute products of features that result in a nonzero value.
This exploits the sparsity of the data to achieve an improved time complexity of $O(d^kD^k)$ on each vector of the matrix where $k$ is the degree of the expansion, $D$ is the dimensionality, and $d$ is the density.
The standard algorithm has time complexity $O(D^k)$.
Since $0 \le d \le 1$, our algorithm is a significant improvement.
While the algorithm we lay out uses CSR matrices, it could be modified to operate on other sparse formats.

@@

For simplicity, all polynomial expansions are assumed to be of the second degree.
We do provide an algorithm for third degree expansions, and derive the big O time complexity of the general case.
We have also developed an algorithm for second and third degree interaction features (combinations without repetition), which can be found in the implementation.

@@

\begin{codebox}
\footnotesize
\Procname{$\proc{Incomplete Second Order CSR Polynomial Expansion Algorithm}(\bm{A})$}
    \li $N \gets$ row count of $\bm{A}$
    \li $\bm{p}^k$ $\gets$ vector of size $N+1$
    \li $\bm{p}^k_0 \gets 0$
    \li $nnz^k \gets 0$
    \li \For $i \gets 0 \To N-1$ \Do
    \li     $i_{start} \gets \bm{p}_i$
    \li     $i_{stop} \gets \bm{p}_{i+1}$
    \li     $\bm{c}_i \gets \bm{c}_{i_{start}:i_{stop}}$
    \li     $nnz^k_i \gets \binom{|\bm{c}_i|}{2}$ \label{li:row_nnz_count}
    \li     $nnz^k \gets nnz^k + nnz^k_i$
    \li     $\bm{p}^k_{i+1} \gets \bm{p}^k_i + nnz^k_i$
        \End
    \zi     
    \zi \Comment Build up the elements of $\bm{p}^k$, $\bm{c}^k$, and $\bm{d}^k$
    \li $\bm{p}^k$ $\gets$ vector of size $N+1$
    \li $\bm{c}^k$ $\gets$ vector of size $nnz^k$
    \li $\bm{d}^k$ $\gets$ vector of size $nnz^k$
    \li $n \gets 0$
    \li \For $i \gets 0 \To N-1$ \Do
    \li     $i_{start} \gets \bm{p}_i$
    \li     $i_{stop} \gets \bm{p}_{i+1}$
    \li     $\bm{c}_i \gets \bm{c}_{i_{start}:i_{stop}}$
    \li     $\bm{d}_i \gets \bm{d}_{i_{start}:i_{stop}}$
    \li     \For $c_1 \gets 0 \To |\bm{c}_i|-1$ \Do
    \li         \For $c_2 \gets c_1 \To |\bm{c}_i|-1$ \Do
%    \li             $j_0 \gets \bm{c}_{c_0}$
%    \li             $j_1 \gets \bm{c}_{c_1}$
    \li             $\bm{d}^k_{n} \gets \bm{d}_{c_0} \cdot \bm{d}_{c_1}$
    \li             $\bm{c}^k_{n} = ?$ \label{li:set_ck}
    \li             $n \gets n + 1$
                \End
            \End
       	\End
\end{codebox}

@@

\begin{equation}
(j_0, j_1, \dots, j_{k-1}) \rightarrowtail \hspace{-1.9ex} \twoheadrightarrow p_{j_0j_1 \dots i_{k-1}} \in \{0,1,\dots,\binom{D}{k}-1\} 
\end{equation}

@@

\section{Final CSR Expansion Algorithm}
With the mapping from columns of $\bm{A}$ to a column of $\bm{A}^k$, we can now write the final form of the innermost loop of the algorithm from \ref{sec:final-algo}.
Let the mapping for $k=2$ be denoted $h^2$.
Then the innermost loop becomes:

@@

The algorithm can be generalized to higher degrees by simply adding more nested loops, using higher order mappings, modifying the output dimensionality, and adjusting the counting of nonzero polynomial features in line \ref{li:row_nnz_count}.
%For interaction features, the interaction mappings can be used in lieu of the polynomial mappings with the additional change of the output dimensionality and the number of nonzero features in each row (combinations without repetition instead of with).

@@

\subsection{Empirical}
To demonstrate how our algorithm scales with the density of a matrix, we compare it to the traditional polynomial expansion algorithm in the popular machine library scikit-learn \cite{scikit-learn} in the task of generating second degree polynomial expansions.
Matrices of size $100 \times 5000$ were randomly generated with densities of $0.2$, $0.4$, $0.6$, $0.8$, and $1.0$.
Thirty matrices of each density were randomly generated, and the mean times (gray) of each algorithm were plotted.
The red or blue width around the mean marks the third standard deviation from the mean.
The time to densify the input to the standard algorithm was not counted.

@@

The fact that our algorithm is approximately $6.5$ times faster than the scikit-learn algorithm on $100 \times 5000$ matrices that are entirely dense is likely a language implementation difference.
What matters is that the time of our algorithm increases quadratically with respect to the density in accordance with the big O analysis.

@@

\section{Conclusion}
We have developed an algorithm for performing polynomial feature expansions on CSR matrices that scales polynomially with respect to the density of the matrix.
While the areas within machine learning this work touches are not en vogue, they are workhorses of industry.
This improvement could therefore spare the burning of much fossil fuel.
